---
title: "Basic R: Matrices"
author: "Jane Briglio"
date: "January 25, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Matrix problems

### 1. Suppose
$$ A = \begin{bmatrix}
1 & 1 & 3\\ 
5 & 2 & 6\\ 
-2 & -1 & -3
\end{bmatrix}$$
(a) Check that $A^3 = \mathbf{0}$  
(b) Replace the third column of A by the 
sum of the second and third columns  
   
part a 
```{r}
A <- matrix(c(1,1,3,5,2,6,-2,-1,-3), nrow = 3, byrow = TRUE)
A
A*A*A
```
part b
```{r}
A[,3] <- A[,2] + A[,3]
A
```

### 2. Create the following matrix B with 15 rows
$$B = \begin{bmatrix}
10 & -10 & 10\\
10 &  -10& 10\\
 \cdots & \cdots  & \cdots \\
10 & -10 & 10
\end{bmatrix} $$
Calculate the 3x3 matrix $B^TB$.  You can make this calculation with the function crossprod().  See the documentaion. 

```{r}
x <- rep(10, times=15)
y <- rep(-10, times=15)
b <- c(x, y, x)
bb <- matrix(b, 15, 3)
bb
crossprod(bb, bb) 
```

### 3. Create a 6 x 6 matrix matE  with every element equal to 0.  check what the functions row() and col() return when applied to matE.
Now, create the 6 x 6 matix:
$$\begin{bmatrix}
0 &1  & 0 & 0 & 0 & 0\\
1 &0 & 1 &0  & 0 &0 \\
0 & 1 & 0 & 1 & 0 & 0\\
0 & 0 & 1 & 0 & 1 & 0\\
0& 0 & 0 & 1 & 0 & 1\\
0& 0 & 0 & 0 & 1 & 0
\end{bmatrix} $$  

```{r}
matE<- matrix(0, 6, 6)
row(matE)
col(matE)
row(matE)-col(matE)
matE[abs(row(matE)-col(matE))==1] <- 1
matE

```

### 4.  Look at the help for the function outer().  Now, create the following patterned matrix:
$$\begin{bmatrix}
0 & 1 & 2 & 3 & 4\\
1 & 2 & 3 & 4 & 5\\
2 & 3 & 4 & 5 & 6\\
3 & 4 & 5 & 6 & 7\\
4 & 5 & 6 & 7 & 8
\end{bmatrix} $$

```{r}
a <- 0:4
A <- outer(a,a,"+")
A
```

### 5.  Create the following patterned matrices.  Your solutions should be generalizable to enable creating larger matrices with the same structure.
(a)
$$\begin{bmatrix}
0 & 1 & 2 & 3 & 4\\
1 &2 & 3 & 4 & 0\\
2 & 3 & 4 & 0 & 1\\
3 & 4 & 0 & 1 & 2\\
4 & 0 & 1 & 2 & 3
\end{bmatrix}  $$
```{r}
x <- outer(0:4, 0:4, "+")
matrix(x%%5, 5, 5)
```
(b)
$$ \begin{bmatrix}
0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9\\
1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 0\\
\vdots  & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
8 & 9 & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7\\
9 & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8
\end{bmatrix} $$
```{r}
x <- outer(0:9, 0:9, "+")
matrix(x%%10, 10, 10)
```
(c)
$$\begin{bmatrix}
0 & 8 & 7 & 6 & 5 & 4 &3  & 2 & 1\\
1 & 0 & 8 & 7 & 6 & 5 & 4 & 3 & 2\\
2 & 1 & 0 & 8 & 7 & 6 & 5 & 4 & 3\\
3 & 2 & 1 & 0 & 8 & 7 & 6 & 5 & 4\\
4 & 3 & 2 & 1 & 0 & 8 & 7 & 6 & 5\\
5 & 4 & 3 & 2 & 1 & 0 & 8 & 7 & 6\\
6 & 5 & 4 & 3 & 2 & 1 & 0 & 8 & 7\\
7 & 6 & 5 & 4 & 3 & 2 & 1 & 0 & 8\\
8 & 7 & 6 & 5 & 4 & 3 & 2 & 1 & 0
\end{bmatrix} $$
```{r}
x <- outer(0:8, 0:8 ,"-")
matrix(x%%9, 9, 9)
```

### 6.  Solve the following system of linear equations by setting up and solving the matrix equation Ax = y.
$x_1 + 2x_2 + 3x_3 + 4x_4 + 5x_5 =  7$  
$2x_1 +  x_2 + 2x_3 + 3x_4 + 4x_5 = -1$  
$3x_1 + 2x_2 +  x_3 + 2x_4 + 3x_5 = -3$  
$4x_1 + 3x_2 + 2x_3 +  x_4 + 2x_5 =  5$  
$5x_1 + 4x_2 + 3x_3 + 2x_4 +  x_5 = 17$ 

```{r}
a <- 1:5
b <- c(2, 1:4)
d <- c(3, 2, 1:3)
e <- c(4:1, 2)
f <- 5:1
A <- matrix(c(a, b, d, e, f), 5, 5)
b <- matrix(c(7, -1, -3, 5, 17), 5, 1)
solve(A,b)
```

### 7. Create a 6 x 10 matrix of random integers chosen from 1,2,...,10 by executing the following two lines of code:

set.seed(75)  
aMat <- matrix(sample(10, size=60, replace=TRUE), nr=6)  

Use the matrix you have created to answer these questions:  
(a) Find the number of entries in each row which are greater than 4.

```{r}
set.seed(75)  
aMat <- matrix(sample(10, size=60, replace=TRUE), nr=6)
length(which(aMat[1,]>4))
length(which(aMat[2,]>4))
length(which(aMat[3,]>4))
length(which(aMat[4,]>4))
length(which(aMat[5,]>4))
length(which(aMat[6,]>4))
```
(b) Which rows contain exactly two occurrences of the number seven?
```{r}
a <- length(which(aMat[1,]==7))
b <- length(which(aMat[2,]==7))
c <- length(which(aMat[3,]==7))
d <- length(which(aMat[4,]==7))
e <- length(which(aMat[5,]==7))
f <- length(which(aMat[6,]==7))
x <- c(a, b, c, d, e, f)
which(x==2)
```
(c) Find those pairs of columns whose total (over both columns) is greater tnan 75.  The answer should be a matrix with two columns; so, for example, the row (1,2) in the output matrix means that the sum of columns 1 and 2 in the original matrix is greater than 75.  Repeating a column is permitted; so, for exalmple, the final output matrix could contain the rows (1,2), (2,1), and (2,2).  
```{r}
x <- colSums(aMat)
which(outer(x,x,"+")>75, arr.ind=T )
```
What if repetitions are not permitted? Then only (1,2) from (1,2),(2,1) and (2,2) would be permitted.
```{r}
x <- colSums(aMat)
y <- outer(x,x,"+")>75
y [lower.tri(y ,diag=T)] <- F
which(y , arr.ind=T)
```

### 8. Calculate
(a) $\sum\limits_{i=1}^{20} \sum\limits_{j=1}^5 \frac{i^4}{(3+j)}$
```{r}
i <- 1:20
j <- 1:5
sum(outer(i^4, (3+(j)), "/"))
```
(b) $\sum\limits_{i=1}^{20} \sum\limits_{j=1}^5 \frac{i^4}{(3+ij)}$
```{r}
i <- 1:20
j <- 1:5
sum((i^4)/ (3 + outer(i,j,"*")))
```
(c) $\sum\limits_{i=1}^{10} \sum\limits_{j=1}^i \frac{i^4}{(3+ij)}$
```{r}
i <- 1:20
j <- 1:5
sum(outer(1:10,1:10,function(i,j){ (i>=j)*i^4/(3+i*j) }) )
```

